Management induced changes of soil organic carbon on global croplands

### 2022-07-20 SOCbudgetPaper/methods.Rmd
update to final version - 2. resubmit

According to the IPCC guidelines $SOC$ changes can be expressed as the difference of two consecutive years [see Eq. 5.0A in @ogle_cropland_in_ipcc_2019]. This, however, will also include naturally occurring changes due to climatic variation over time. For our study, we define the absolute and relative $SOC$ changes in relation to a potential natural state $SOC^{\mathrm{pnv}}$ under the same climatic conditions in grid cell $i$ at time $t$, that is based on the natural vegetation $SOC$ calculations as defined above without accounting for land conversion from cropland at any time. The absolute changes $\Delta SOC$ and relative changes $F^{\mathrm{SCF}}$ are thus given by

@@

From our 17 crop groups, we compile four residue groups (straw, high- and low-lignin residues, residues without dual use), of which the first three are taken away from the field for other purposes (see mappingCrop2Residue.csv in Bodirsky et al. [-@bodirsky_mrcommons_2020]). Residue feed demand for five different livestock groups is based on country-specific feed baskets [see @weindl_livestock_2017], that differentiate between the residue groups and take available $AGR$ biomass as well as livestock productivity into account. We estimate a material-use share for the straw-residue group of 5% and a fuel-share of 10% for all used residue groups in low-income countries. For high-income countries, no withdrawal for material or fuel use is assumed, and use shares of middle-income countries are linearly interpolated based on per-capita income, following the same rationale as for the share of burnt residues described above. The remaining $AGR$ as well as all $BGR$ are expected to be left on the field. We limit high residue return rates to at most $10\,\unit{tC\,ha}^{-1}$ in order to correct for outliers.

@@

Manure production and management is based on a revised methodology of Bodirsky et al. [-@bodirsky_n2o_2012] and is presented here due to its central role in soil carbon modeling. Based on the gridded livestock distribution we calculate spatially explicit excretion by estimating the nitrogen balance of the livestock systems on the basis of comprehensive livestock feed baskets [@weindl_livestock_2017], assuming that all nitrogen in protein feed intake, minus the nitrogen in the slaughter mass, is excreted. Carbon in excreted manure is estimated by applying fixed C:N ratios, which range from 10 for poultry up to 19 for beef cattle (for full detail see Calvo Buendia et al. [-@ogle_cropland_in_ipcc_2019]. 
Depending on the feed system we assume manure to be handled in four different ways:
All manure originated from pasture feed intake is excreted directly on pastures and rangelands (pasture grazing), deducting manure collected as fuel.
Whereas for low-income countries, we adopt a share of 25% of crop residues in feed intake directly consumed and excreted on crop fields (stubble grazing), we do not consider any stubble grazing in high-income countries; middle-income countries see linearly interpolated shares depending on their per-capita income.
For all other feed items, we assume the manure to be stored in animal waste management systems associated with livestock housing. 
To estimate the carbon actually returned to the soil, we account for carbon losses during storage, where return shares depend on different animal waste management and grazing systems. Whereas we assume no losses for pasture and stubble grazing, we consider that the manure collected as fuel is not returned to the fields. For manure stored in different animal waste management systems we compiled carbon loss rates (see calcClossConfinement.R in Bodirsky et al. [-@bodirsky_mrcommons_2020] for more details) depending on the different systems and the associated nitrogen loss rates as specified in Bodirsky et al. [-@bodirsky_n2o_2012]. We limit high application shares at $10\,\unit{tC\,ha}^{-1}$ to correct for outliers, that can occur due to inconsistencies between FAO production and 0.5 degree land-use data.

>>>>

According to the IPCC guidelines $SOC$ changes can be expressed as the difference of two consecutive years [see Eq. 5.0A in @ogle_cropland_in_ipcc_2019]. This, however, will also include naturally occurring changes due to climatic variation over time. For our study, we define the absolute and relative $SOC$ changes in relation to a potential natural state $SOC^{\mathrm{pnv}}$ under the same climatic conditions in grid cell $i$ at time $t$ that is based on the natural vegetation $SOC$ calculations as defined above without accounting for land conversion from cropland at any time. The absolute changes $\Delta SOC$ and relative changes $F^{\mathrm{SCF}}$ are thus given by

@@

From our 17 crop groups, we compile four residue groups (straw, high- and low-lignin residues, residues without dual use), of which the first three are taken away from the field for other purposes (see mappingCrop2Residue.csv in Bodirsky et al. [-@bodirsky_mrcommons_2020]). Residue feed demand for five different livestock groups is based on country-specific feed baskets [see @weindl_livestock_2017] that differentiate between the residue groups and take available $AGR$ biomass as well as livestock productivity into account. We estimate a material-use share for the straw-residue group of 5% and a fuel-share of 10% for all used residue groups in low-income countries. For high-income countries, no withdrawal for material or fuel use is assumed, and use shares of middle-income countries are linearly interpolated based on per-capita income, following the same rationale as for the share of burnt residues described above. The remaining $AGR$ as well as all $BGR$ are expected to be left on the field. We limit high residue return rates to at most $10\,\unit{tC\,ha}^{-1}$ in order to correct for outliers.

@@

Manure production and management is based on a revised methodology of Bodirsky et al. [-@bodirsky_n2o_2012] and is presented here due to its central role in soil carbon modeling. Based on the gridded livestock distribution we calculate spatially explicit excretion by estimating the nitrogen balance of the livestock systems on the basis of comprehensive livestock feed baskets [@weindl_livestock_2017], assuming that all nitrogen in protein feed intake, minus the nitrogen in the slaughter mass, is excreted. Carbon in excreted manure is estimated by applying fixed C:N ratios, which range from 10 for poultry up to 19 for beef cattle (for full detail see Calvo Buendia et al. [-@ogle_cropland_in_ipcc_2019]. 
Depending on the feed system we assume manure to be handled in four different ways:
All manure originated from pasture feed intake is excreted directly on pastures and rangelands (pasture grazing), deducting manure collected as fuel.
Whereas for low-income countries, we adopt a share of 25% of crop residues in feed intake directly consumed and excreted on crop fields (stubble grazing), we do not consider any stubble grazing in high-income countries; middle-income countries see linearly interpolated shares depending on their per-capita income.
For all other feed items, we assume the manure to be stored in animal waste management systems associated with livestock housing. 
To estimate the carbon actually returned to the soil, we account for carbon losses during storage, where return shares depend on different animal waste management and grazing systems. Whereas we assume no losses for pasture and stubble grazing, we consider that the manure collected as fuel is not returned to the fields. For manure stored in different animal waste management systems we compiled carbon loss rates (see calcClossConfinement.R in Bodirsky et al. [-@bodirsky_mrcommons_2020] for more details) depending on the different systems and the associated nitrogen loss rates as specified in Bodirsky et al. [-@bodirsky_n2o_2012]. We limit high application shares at $10\,\unit{tC\,ha}^{-1}$ to correct for outliers that can occur due to inconsistencies between FAO production and 0.5 degree land-use data.

### 2022-07-20 SOCbudgetPaper/results.Rmd
update to final version - 2. resubmit

In Fig. \@ref(fig:SOCmaps)(b) we provide a world map of SOC stocks estimates for the first $30\,\unit{cm}$ on cropland considering historical management data for the year 2010. Values range between over $100\,\unit{t\,ha^{-1}}$ in northern temperate cropland to less than $5\,\unit{t\,ha^{-1}}$ for arid and semiarid cropland.
Our spatially explicit results show hotspots of SOC losses and gains compared to SOC under PNV in two complementary ways: 
1. Absolute SOC changes $\Delta SOC$ (Fig. \@ref(fig:SOCmaps)(c)) indicate areas with high importance for the global $SOC$ loss. They can be driven by large relative changes (e.g. in Central Africa) or by a high natural stock, from which even small relative deviations could lead to substantial absolute losses (e.g. North-East Asia). 
2. Relative SOC changes measured as stock changes factors $F^{SCF}$ (Fig. \@ref(fig:SOCmaps)(d)) are a helpful metric to analyze the impact of human cropping activities. They indicate areas with large differences in carbon inflows or SOC decay compared to natural vegetation, that may hold potential to be overcome by improved agricultural practices. Large parts of tropical cropland seem to suffer from strongly reduced relative stocks, indicating SOC degradation. Conversely, irrigated cropland at the border to dry, unsuitable areas worldwide shows a strong relative increase in SOC stocks.

@@

Using the constManagement results, that only include land-use change (LUC) related changes of the SOC debt between 1975 and 2010, we are able to subtract the LUC effect from the overall SOC debt change within the histManagement results. The remaining effect can be attributed to the changing agricultural management (MAN) as other drivers such as climatic effects have been already canceled out by taking the difference to a PNV reference state while calculating $\Delta SOC$. 
The increasing SOC debt on global cropland are primarily caused by LUC (red areas in Fig. \@ref(fig:SOCdebt)(b2)). Deteriorated management also contributed to increasing SOC debt in parts of Sub-saharan Africa and Central Asia. In contrast, agricultural management has led to an decrease in SOC debt in the USA, Europe, as well as in parts of China and India (blue areas in Fig. \@ref(fig:SOCdebt)(b3)), which is not visible in the total $\Delta SOC$ change as LUC was happening at the same time.

@@

To evaluate our modeled SOC stocks and stock changes under agricultural management, we compare our results to the default IPCC stock change factors $F^{\mathrm{SCF}}$ of 2006 [@lasco_cropland_in_ipcc_2006] and their refinements in 2019 [@ogle_cropland_in_ipcc_2019]. Both estimates are based on measurement data for cropland (see Fig. \@ref(fig:SCFglo)). To allow for comparison, we aggregate our stock change factors weighted by grid-level cropland area to derive median factors for the four IPCC climate zones (Fig. \@ref(fig:CLIMzone)). Note that IPCC Tier 1 factors are derived under the assumption that there is a linear change between steady states over 20 years, whereas our aggregated factors just reflect the relative change compared to a given potential natural vegetation reference stock without specifically tracking the time scale of the stock change.

@@

```{r SCFglo, out.width = "15cm", fig.cap = captionSCFglo, echo=FALSE, fig.pos="h"}
# All defaults
knitr::include_graphics("../ResultNotebooks/Output/Images/SCF_comparison.png")
```

@@

The global estimates of the total SOC stock of the upper 30 cm from this study are in the middle of the wide range of other modeled or observation-based estimates. Regional results (Fig. \@ref(fig:SOCreg)) show that our estimates are well within the range of other estimates for most regions, but at the lower end for tropical moist and tropical wet areas. SoilGrids [@hengl_soilgrids250m_2017] especially stands out with its high estimate, since they include the litter horizon on top of the soil, that might dominate especially polar and boreal soils. SoilGrids 2.0 [@poggio_soilgrids_2021] however, excludes litter C and thus marks the lower end for particularly northern regions. For the same reason it is also more comparable to our results, which do not account for litter C as well.

@@

Estimates of SOC stocks under natural vegetation influence our modeled results for cropland, which has been converted from natural vegetation at some point in time. As the Tier 2 modeling approach [@ogle_cropland_in_ipcc_2019] is not specifically parameterized for natural vegetation it is important to evaluate its suitability to produce reasonable results in that domain as well at least comparable to other modeling approaches.
We therefore also compare our modeled results for SOC under natural vegetation (derived using litterfall of LPJmL4) against estimates of SOC by LPJmL4 for a PNV simulation (see Fig.\@ref(fig:SOCLPJmL4)). Both models are driven by the same climate conditions and the same natural litterfall and just differ in the representation of SOC and litter dynamics. With our focus on cropland SOC dynamics, we compare only cells with more than $1000\,\unit{ha}$ of cropland (capturing 99.9% of global cropland area).
Spatial correlations of PNV SOC stock values are high (global $R2=0.81$), especially for dry climate zones (Fig. \@ref(fig:SOCLPJmL4)). For temperate and tropical moist areas estimates of this study tend to be a bit lower compared to LPJmL4 results.

>>>>

In Fig. \@ref(fig:SOCmaps)(b) we provide a world map of SOC stocks estimates for the first $30\,\unit{cm}$ on cropland considering historical management data for the year 2010. Values range between over $100\,\unit{t\,ha^{-1}}$ in northern temperate cropland to less than $5\,\unit{t\,ha^{-1}}$ for arid and semiarid cropland.
Our spatially explicit results show hotspots of SOC losses and gains compared to SOC under PNV in two complementary ways: 
1. Absolute SOC changes $\Delta SOC$ (Fig. \@ref(fig:SOCmaps)(c)) indicate areas with high importance for the global $SOC$ loss. They can be driven by large relative changes (e.g. in Central Africa) or by a high natural stock, from which even small relative deviations could lead to substantial absolute losses (e.g. North-East Asia). 
2. Relative SOC changes measured as stock changes factors $F^{SCF}$ (Fig. \@ref(fig:SOCmaps)(d)) are a helpful metric to analyze the impact of human cropping activities. They indicate areas with large differences in carbon inflows or SOC decay compared to natural vegetation that may hold potential to be overcome by improved agricultural practices. Large parts of tropical cropland seem to suffer from strongly reduced relative stocks, indicating SOC degradation. Conversely, irrigated cropland at the border to dry, unsuitable areas worldwide shows a strong relative increase in SOC stocks.

@@

Using the constManagement results that only include land-use change (LUC) related changes of the SOC debt between 1975 and 2010, we are able to subtract the LUC effect from the overall SOC debt change within the histManagement results. The remaining effect can be attributed to the changing agricultural management (MAN) as other drivers such as climatic effects have been already canceled out by taking the difference to a PNV reference state when calculating $\Delta SOC$. 
The increasing SOC debt on global cropland are primarily caused by LUC (red areas in Fig. \@ref(fig:SOCdebt)(b2)). Deteriorated management also contributed to increasing SOC debt in parts of Sub-saharan Africa and Central Asia. In contrast, agricultural management has led to an decrease in SOC debt in the USA, Europe, as well as in parts of China and India (blue areas in Fig. \@ref(fig:SOCdebt)(b3)), which is not visible in the total $\Delta SOC$ change as LUC was happening at the same time.

@@

To evaluate our modeled SOC stocks and stock changes under agricultural management, we compare our results to the default IPCC stock change factors $F^{\mathrm{SCF}}$ of 2006 [@lasco_cropland_in_ipcc_2006] and their refinements in 2019 [@ogle_cropland_in_ipcc_2019]. Both estimates are based on measurement data for cropland (see Fig. \@ref(fig:SCFglo)). To allow for comparison, we aggregate our stock change factors weighted by grid-level cropland area to derive median factors for the four IPCC climate zones (Fig. \@ref(fig:CLIMzone)). Note that IPCC Tier 1 factors are derived under the assumption that there is a linear change between steady states over 20 years, whereas our aggregated factors just reflect the relative change compared to a given potential natural vegetation reference stock without specifically tracking age of the cropland.

@@

```{r SCFglo, out.width = "16cm", fig.cap = captionSCFglo, echo=FALSE, fig.pos="h"}
# All defaults
knitr::include_graphics("../ResultNotebooks/Output/Images/SCF_comparison.png")
```

@@

The global estimates of the total SOC stock of the upper 30 cm from this study are in the middle of the wide range of other modeled or observation-based estimates. Regional results (Fig. \@ref(fig:SOCreg)) show that our estimates are well within the range of other estimates for most regions, but at the lower end for tropical moist and tropical wet areas. SoilGrids [@hengl_soilgrids250m_2017] especially stands out with its high estimate, since they include the litter horizon on top of the soil that might dominate especially polar and boreal soils. SoilGrids 2.0 [@poggio_soilgrids_2021] however, excludes litter C and thus marks the lower end for particularly northern regions. For the same reason it is also more comparable to our results, which do not account for litter C as well.

@@

Estimates of SOC stocks under natural vegetation influence our modeled results for cropland, which has been converted from natural vegetation at some point in time. As the Tier 2 modeling approach [@ogle_cropland_in_ipcc_2019] is not specifically parameterized for natural vegetation it is important to evaluate its suitability to produce reasonable results in that domain as well at least comparable to other modeling approaches.
We therefore also compare our modeled results for SOC under natural vegetation (derived using litterfall of LPJmL4) against estimates of SOC by LPJmL4 for a PNV simulation (see Fig. \@ref(fig:SOCLPJmL4)). Both models are driven by the same climate conditions and the same natural litterfall and just differ in the representation of SOC and litter dynamics. With our focus on cropland SOC dynamics, we compare only cells with more than $1000\,\unit{ha}$ of cropland (capturing 99.9% of global cropland area).
Spatial correlations of PNV SOC stock values are high (global $R2=0.81$), especially for dry climate zones (Fig. \@ref(fig:SOCLPJmL4)). For temperate and tropical moist areas estimates of this study tend to be a bit lower compared to LPJmL4 results.
